---
title: "`summarize` and the pipe operator (`%>%`)"
date: 'Last compiled: `r format(Sys.time(), "%b %d, %Y")`'
author: "Your Name Here"
output:
  html_document:
    df_print: paged
  pdf_document: default
---

```{r, label = "SETUP", echo = FALSE, results= 'hide', message = FALSE, warning = FALSE}
library(knitr)
knitr::opts_chunk$set(echo = TRUE, comment = NA, fig.align = "center", fig.width = 4, fig.height = 4, message = FALSE, warning = FALSE)
library(dplyr)
library(hflights)
library(scales)
# Convert the hflights data.frame into a hflights tbl
hflights <- tbl_df(hflights)
# Create a random subset of size 10,000 to work with
set.seed(123)
train <- sample(227496, 10000)
hflights <- hflights[train, ]
# Create the object carriers, containing only the UniqueCarrier variable of hflights
carriers <- hflights$UniqueCarrier
# Both the dplyr and hflights packages are loaded into workspace
lut <- c("AA" = "American", "AS" = "Alaska", "B6" = "JetBlue", "CO" = "Continental", 
         "DL" = "Delta", "OO" = "SkyWest", "UA" = "United", "US" = "US_Airways", 
         "WN" = "Southwest", "EV" = "Atlantic_Southeast", "F9" = "Frontier", 
         "FL" = "AirTran", "MQ" = "American_Eagle", "XE" = "ExpressJet", 
         "YV" = "Mesa")

# Use lut to translate the UniqueCarrier column of hflights
hflights$UniqueCarrier <- lut[hflights$UniqueCarrier]
lut <- c("A" = "carrier", "B" = "weather", "C" = "FFA", 
         "D" = "security", "E" = "not cancelled")

# Use the lookup table to create a vector of code labels. 
# Assign the vector to the CancellationCode column of hflights
hflights$CancellationCode <- lut[hflights$CancellationCode]
```

_________________

## The syntax of summarize

`summarize()`, the last of the 5 verbs, follows the same syntax as `mutate()`, but the resulting dataset consists of a single row instead of an entire new column in the case of `mutate()`. Below, a typical `summarize()` function is repeated to show the syntax, without going into detail on all arguments:

```{r, eval = FALSE}
summarize(data, sumvar = sum(A), avgvar = avg(B))
```

In contrast to the four other data manipulation functions, `summarize()` does not return a copy of the dataset it is summarizing; instead, it builds a new dataset that contains only the summarzing statistics.

## Instructions

* Use `summarize()` to further explore `hflights`. Determine the shortest distance flown (save this as a variable named `min_dist`) and the longest distance flown (save this as a variable named `max_dist`). Save your solution to variable `s1`.


```{r}
# Your code goes here
# Determine the shortest and longest distance flown and save statistics to 
# min_dist and max_dist resp.


```

* Calculate the longest `Distance` for diverted flights, and save this as a variable named `max_div`. You will need one of the four other verbs to do this! Do this in a one-liner and save the result to `s2`.

```{r}
# Your code goes here
# Determine the longest distance for diverted flights, save statistic to max_div. 
# Use a one-liner!


```

## Aggregate functions

You can use any function you like in `summarize()`, so long as the function can take a vector of data and return a single number. `R` contains many aggregating functions, as `dplyr` calls them. Here are some of the most useful:

* `min(x)` - minimum value of vector `x`.
* `max(x)` - maximum value of vector `x`.
* `mean(x)` - mean value of vector `x`.
* `median(x)` - median value of vector `x`.
* `quantile(x, p)` - pth quantile of vector `x`.
* `sd(x)` - standard deviation of vector `x`.
* `var(x)` - variance of vector `x`.
* `IQR(x)` - Inter Quartile Range (IQR) of vector `x`.
* `diff(range(x))` - total range of vector `x`.

## Instructions

* Remove rows that have `NA`'s in the arrival delay column and save the resulting dataset to `temp1`. Then create a table with the following variables (and variable names): the minimum arrival delay (`earliest`), the average arrival delay (`average`), the longest arrival delay (`latest`), and the standard deviation for arrival delays (`sd`). Save this `data.frame` in the variable `s1`.

```{r}
# Your code goes here


```

Filter `hflights` such that only rows that have no `NA` `TaxiIn` and no `NA` `TaxiOut` are kept; save this temporary result to `temp2`. Then create a `data.frame` using `summarize()`, with one variable, `max_taxi_diff` that shows the biggest absolute difference in time between `TaxiIn` and `TaxiOut` for a single flight.

```{r}
# Your code goes here
# Calculate the maximum taxiing difference for flights that have taxi data available


```

## `dplyr` aggregate functions

`dplyr` provides several helpful aggregate functions of its own, in addition to the ones that are already defined in `R`. These include:

* `first(x)` - The first element of vector `x`.
* `last(x)` - The last element of vector `x`.
* `nth(x, n)` - The nth element of vector `x`.
* `n()` - The number of rows in the data.frame or group of observations that `summarize()` describes.
* `n_distinct(x)` - The number of unique values in vector `x`.

Next to these dplyr-specific functions, you can also turn a logical test into an aggregating function with `sum()` or `mean()`. A logical test returns a vector of `TRUE`'s and `FALSE`'s. When you apply `sum()` or `mean()` to such a vector, `R` coerces each `TRUE` to a 1 and each `FALSE` to a 0. This allows you to find the total number or proportion of observations that passed the test, respectively.

## Instructions

* Create a table with the following variables (and variable names): the total number of observations in `hflights` (`n_obs`), the total number of carriers that appear in `hflights` (`n_carrier`), the total number of destinations that appear in `hflights` (`n_dest`), and the destination of the flight that appears in the 100th row of `hflights` (`dest100`). Save the result to `s1`.

```{r}
# Your code goes here
# Calculate the summarizing statistics of hflights


```

* Keep all of the flights in `hflights` flown by American Airlines ("American"), and assign the result to `aa`. Then calculate the total number of flights flown by American Airlines (`n_flights`), the total number of cancelled flights (`n_canc`), the percentage of cancelled flights (`p_canc`), and the average arrival delay of the flights whose delay does not equal `NA` (`avg_delay`). Store the final result in `s2`.

```{r}
# Your code goes here
# Calculate the summarizing statistics for flights flown by 
# American Airlines (carrier code "American")


```

## Overview of syntax

The video explained the use and functionality of the `%>%`, but let's make sure you got the point. The following two statements are completely analogous:

```{r, eval = FALSE}
mean(c(1, 2, 3, NA), na.rm = TRUE)
c(1, 2, 3, NA) %>% mean(na.rm = TRUE)
```

The `%>%` operator allows you to extract the first argument of a function from the arguments list and put it in front of it, thus solving the Dagwood sandwich problem.

## Instructions

Use `dplyr` functions and the pipe operator to transform the following English sentences into `R` code:

* Take the `hflights` data set and then …
* Add a variable named `diff` that is the result of subtracting `TaxiIn` from `TaxiOut`, and then …
* Pick all of the rows whose `diff` value does not equal `NA`, and then …
* Summarize the data set with a value named `avg` that is the mean `diff` value.

Store the result in the variable `p`.

```{r}
# Write the 'piped' version of the English sentences.


```

## Drive or fly? Part 1 of 2

You can answer sophisticated questions by combining the verbs of `dplyr`. Over the next few exercises you will examine whether it sometimes makes sense to drive instead of fly. You will begin by making a data set that contains relevant variables. Then, you will find flights whose equivalent average velocity is lower than the velocity when traveling by car.

## Instructions

* Define a data set named `d` that contains just the `Dest`, `UniqueCarrier`, `Distance`, and `ActualElapsedTime` columns of `hflights` as well as two additional variables: `RealTime` and `mph.RealTime` should equal the actual elapsed time plus 100 minutes. This will be an estimate of how much time a person spends getting from point A to point B while flying, including getting to the airport, security checks, etc. `mph` will be the miles per hour that a person on the flight traveled based on the `RealTime` of the flight.

```{r}
# Your code goes here
# Part 1, concerning the selection and creation of columns


```

* On many highways you can drive at 70 mph. Continue with `d` to calculate the following variables: `n_less`, the number of flights whose `mph` value does not equal `NA` that traveled at less than 70 mph in real time; `n_dest`, the number of destinations that were traveled to at less than 70 mph; `min_dist`, the minimum distance of these flights; `max_dist`, the maximum distance of these flights.

```{r}
# Your code goes here
# Part 2, concerning flights that had an actual average speed of < 70 mph.


```

## Drive or fly? Part 2 of 2

The previous exercise suggested that some flights might be less efficient than driving in terms of speed. But is speed all that matters? Flying imposes burdens on a traveler that driving does not. For example, airplane tickets are very expensive. Air travelers also need to limit what they bring on their trip and arrange for a pick up or a drop off. Given these burdens we might demand that a flight provide a large speed advantage over driving.

## Instructions

Let's define preferable flights as flights that are 150% faster than driving, i.e. that travel 105 mph or greater in real time. Also, assume that cancelled or diverted flights are less preferable than driving.

Write an adapted version of the solution to the previous exercise in an all-in-one fashion (i.e. in a single expression without intermediary variables, using `%>%`) to find:

* `n_non` - the number of non-preferable flights in `hflights`,
* `p_non` - the percentage of non-preferable flights in `hflights`,
* `n_dest` - the number of destinations that non-preferable flights traveled to,
* `min_dist` - the minimum distance that non-preferable flights traveled,
* `max_dist` - the maximum distance that non-preferable flights traveled.

To maintain readability in this advanced exercise, start your operations with a `select()` function to retain only the five columns that will be needed for the subsequent calculation steps.

```{r}
# Your code goes here


```

## Advanced piping exercises

Because piping is a very powerful and commonly used feature, you are challenged with one more exercise. Become a piping master!

## Instructions

How many flights were overnight flights (flights whose arrival time is earlier than their departed time). Do not count `NA` values! Use `summarize()` with the `n()` function. The resulting data frame should contain a column named `n`.

```{r}
# Your code goes here
# Count the number of overnight flights


```

________________
